\textbf{The Heart can Lie: A Preliminary Investigation of the Role of Interoception and Theory of Mind in Deception}

### 2024-07-02 paper/manuscript.tex
finalize R2

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,mask,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\linenumbers
\usepackage{csquotes}
\usepackage[labelfont=bf, font={color=gray,small}]{caption}
\usepackage{float}
\usepackage[document]{ragged2e}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The Heart can Lie: A Preliminary Investigation of the Role of Interoception and Theory of Mind in Deception},
  pdflang={en-EN},
  pdfkeywords={Deception; Interoception; Theory of Mind; Polygraph; Lying Ability},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{participants}{%
\subsection{Participants}\label{participants}}

@@

\hypertarget{measures}{%
\subsection{Measures}\label{measures}}

\hypertarget{theory-of-mind-tom}{%
\subsubsection{Theory of Mind (ToM)}\label{theory-of-mind-tom}}

@@

\hypertarget{interoception}{%
\subsubsection{Interoception}\label{interoception}}

@@

\hypertarget{deception}{%
\subsubsection{Deception}\label{deception}}

@@

\hypertarget{procedure}{%
\subsection{Procedure}\label{procedure}}

@@

\hypertarget{data-analysis}{%
\subsection{Data Analysis}\label{data-analysis}}

@@

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{manipulation-check}{%
\subsection{Manipulation Check}\label{manipulation-check}}

@@

\hypertarget{feature-reduction}{%
\subsection{Feature Reduction}\label{feature-reduction}}

@@

\hypertarget{theory-of-mind}{%
\subsection{Theory of Mind}\label{theory-of-mind}}

@@

\hypertarget{interoception-1}{%
\subsection{Interoception}\label{interoception-1}}

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

The present study examined the contribution of ToM and interoception on our ability to lie using a directed lie paradigm with two conditions (``Interrogation'' and ``Polygraph'') designed to enhance each of the two mechanisms. Interestingly, we found that when participants were presented with (fake) physiological feedback (the polygraph condition), instead of a face of a person they had to lie to (the interrogation condition), their response time for both lies and truths increased, as did their heart rate. Although the condition did not impact the subjective confidence that participants had in their answers, it suggests that believing oneself to be submitted to a machine that is supposedly able to detect deception by interpreting physiological signals is a harder and/or more stressful condition than lying to a person. While research linking interoception and deception is limited, our results are in line with studies that show an association between interoceptive awareness and anxiety (Domschke et al., 2010; Garfinkel \& Critchley, 2013; Yoris et al., 2015). Specifically, enhancing one's attention towards their internal bodily signals could have resulted in a hyper-vigilance towards physiological sensations that is perceived negatively. This is consistent with previous deception detection studies, in which participants' spontaneous lying behaviour only decreased when they were given feedback by a polygraph machine, but not when no feedback was given (Peleg et al., 2019). By extension, our study adds to the controversial discourse surrounding the use of physiological measures in past deception research, further questioning its validity as an indicator of deception (Oviatt et al., 2018; Rosky, 2013).

Our results suggest that higher ToM abilities were related to slower and less confident lies, but only in the polygraph condition. While previous bodies of work have reported mixed findings regarding the association between interoception and ToM (Chiou \& Lee, 2013; Gendolla \& Wicklund, 2009; Scaffidi Abbate et al., 2016; Wundrack \& Specht, 2023), our results suggest the two are negatively linked. One possible interpretation of our findings is that people with stronger ToM abilities by default rely more on their social skills and altercentric inference when lying (i.e., they focus on - and try to read - the other person). When that mechanism is unavailable or unsuited (e.g., when there is no person to lie to - but a ``machine'' in our case), their corresponding lying ability decreases. However, in light of the current field of mixed findings relating interoception and ToM (Canino et al., 2022; Gao et al., 2019; Miller, 2015; Shah et al., 2017), future studies are necessary to investigate the interaction of these mechanisms in different social contexts.

We also showed that interoceptive abilities (as indicated by the composite interoception scores) are correlated with a higher confidence in one's lies in the polygraph condition, a condition in which the attention towards internal reactions is fostered. Indeed, this is in line with previous studies that found individuals with low interoception were more averse to risk when reputational stakes were high, telling fewer egoistical lies (Vabba et al., 2022). In fact, Vabba et al. (2022) further reported that people with high interoception abilities were less likely to differ in risk-taking tendencies, telling the same number of lies regardless of the social stakes. Consistent with our results, Mohr et al. (2023) found that individuals with high interoceptive accuracy were more likely to make egocentric decisions. However, in contrast to previous studies (Füstös et al., 2013; Owens et al., 2018; Pinna \& Edwards, 2020; Pollatos et al., 2007), we did not find any significant relationship between individuals' interoception scores and their heart rate changes during their answers. This points toward a predominantly meta-cognitive effect without necessarily an actual bodily regulation (i.e., participants with good interoception feel that their lies are more convincing, but do not actively attenuate their bodily reactions).

@@

Additionally to the limitations pertaining to the measure of lying ability, some also concern the measure of the predictor constructs, namely ToM and interoception. While we tried to include a behavioral task as well as a subjective questionnaire for each, it has to be underlined that they are notoriously difficult concepts to measure. In particular, objective interoceptive accuracy was assessed using the Heartbeat Counting Task (HCT). While the HCT used to be considered as a gold standard and remains one of the most commonly used measures (Desmedt et al., 2022), concerns regarding its validity have been highlighted in several studies (Brener \& Ring, 2016; Desmedt et al., 2018, 2022; Legrand et al., 2022). Given increasing research efforts invested in the development of novel interoception tasks (Legrand et al., 2022; Plans et al., 2021; Ponzo et al., 2021), future works should further examine the relationship between interoception and lying ability using measures with better psychometric properties.

@@

\hypertarget{data-availability}{%
\section{Data Availability}\label{data-availability}}

@@

\hypertarget{conflict-of-interest-statement}{%
\section{Conflict of Interest Statement}\label{conflict-of-interest-statement}}

@@

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

@@

\hypertarget{references}{%
\section{References}\label{references}}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-abe2011brain}{}}%
Abe, N. (2011). How the brain shapes deception: An integrated review of the literature. \emph{The Neuroscientist}, \emph{17}(5), 560--574.

\leavevmode\vadjust pre{\hypertarget{ref-abe2007deceiving}{}}%
Abe, N., Suzuki, M., Mori, E., Itoh, M., \& Fujii, T. (2007). Deceiving others: Distinct neural responses of the prefrontal cortex and amygdala in simple fabrication and deception with social interactions. \emph{Journal of Cognitive Neuroscience}, \emph{19}(2), 287--295.

\leavevmode\vadjust pre{\hypertarget{ref-baron1997mindblindness}{}}%
Baron-Cohen, S. (1997). \emph{Mindblindness: An essay on autism and theory of mind}. MIT press.

\leavevmode\vadjust pre{\hypertarget{ref-barrett2015interoceptive}{}}%
Barrett, L. F., \& Simmons, W. K. (2015). Interoceptive predictions in the brain. \emph{Nature Reviews Neuroscience}, \emph{16}(7), 419--429.

\leavevmode\vadjust pre{\hypertarget{ref-battista2021role}{}}%
Battista, F., Otgaar, H., Mangiulli, I., \& Curci, A. (2021). The role of executive functions in the effects of lying on memory. \emph{Acta Psychologica}, \emph{215}, 103295.

\leavevmode\vadjust pre{\hypertarget{ref-baumgartner2013honest}{}}%
Baumgartner, T., Gianotti, L. R., \& Knoch, D. (2013). Who is honest and why: Baseline activation in anterior insula predicts inter-individual differences in deceptive behavior. \emph{Biological Psychology}, \emph{94}(1), 192--197.

\leavevmode\vadjust pre{\hypertarget{ref-beaudoin2020systematic}{}}%
Beaudoin, C., Leblanc, É., Gagner, C., \& Beauchamp, M. H. (2020). Systematic review and inventory of theory of mind measures for young children. \emph{Frontiers in Psychology}, \emph{10}, 2905.

\leavevmode\vadjust pre{\hypertarget{ref-bond2008individual}{}}%
Bond Jr, C. F., \& DePaulo, B. M. (2008). Individual differences in judging deception: Accuracy and bias. \emph{Psychological Bulletin}, \emph{134}(4), 477.

\leavevmode\vadjust pre{\hypertarget{ref-bora2017meta}{}}%
Bora, E., \& Yener, G. G. (2017). Meta-analysis of social cognition in mild cognitive impairment. \emph{Journal of Geriatric Psychiatry and Neurology}, \emph{30}(4), 206--213.

\leavevmode\vadjust pre{\hypertarget{ref-brener2016towards}{}}%
Brener, J., \& Ring, C. (2016). Towards a psychophysics of interoceptive processes: The measurement of heartbeat detection. \emph{Philosophical Transactions of the Royal Society B: Biological Sciences}, \emph{371}(1708), 20160015.

\leavevmode\vadjust pre{\hypertarget{ref-burgoon1994interpersonal}{}}%
Burgoon, J. K., \& Buller, D. B. (1994). Interpersonal deception: III. Effects of deceit on perceived communication and nonverbal behavior dynamics. \emph{Journal of Nonverbal Behavior}, \emph{18}(2), 155--184.

\leavevmode\vadjust pre{\hypertarget{ref-Burkner2017}{}}%
Bürkner, P.-C. (2017). {brms}: An {R} package for {Bayesian} multilevel models using {Stan}. \emph{Journal of Statistical Software}, \emph{80}(1), 1--28. \url{https://doi.org/10.18637/jss.v080.i01}

\leavevmode\vadjust pre{\hypertarget{ref-canino2022embodiment}{}}%
Canino, S., Raimo, S., Boccia, M., Di Vita, A., \& Palermo, L. (2022). On the embodiment of social cognition skills: The inner and outer body processing differently contributes to the affective and cognitive theory of mind. \emph{Brain Sciences}, \emph{12}(11), 1423.

\leavevmode\vadjust pre{\hypertarget{ref-chen2021emerging}{}}%
Chen, W. G., Schloesser, D., Arensdorf, A. M., Simmons, J. M., Cui, C., Valentino, R., Gnadt, J. W., Nielsen, L., Hillaire-Clarke, C. S., Spruance, V., et al. (2021). The emerging science of interoception: Sensing, integrating, interpreting, and regulating signals within the self. \emph{Trends in Neurosciences}, \emph{44}(1), 3--16.

\leavevmode\vadjust pre{\hypertarget{ref-chiou2013enactment}{}}%
Chiou, W.-B., \& Lee, C.-C. (2013). Enactment of one-to-many communication may induce self-focused attention that leads to diminished perspective taking: The case of facebook. \emph{Judgment and Decision Making}, \emph{8}(3), 372--380.

\leavevmode\vadjust pre{\hypertarget{ref-connell2018interoception}{}}%
Connell, L., Lynott, D., \& Banks, B. (2018). Interoception: The forgotten modality in perceptual grounding of abstract and concrete concepts. \emph{Philosophical Transactions of the Royal Society B: Biological Sciences}, \emph{373}(1752), 20170143.

\leavevmode\vadjust pre{\hypertarget{ref-craig2009you}{}}%
Craig, A. D. (2009). How do you feel---now? The anterior insula and human awareness. \emph{Nature Reviews Neuroscience}, \emph{10}(1), 59--70.

\leavevmode\vadjust pre{\hypertarget{ref-critchley2004neural}{}}%
Critchley, H. D., Wiens, S., Rotshtein, P., Öhman, A., \& Dolan, R. J. (2004). Neural systems supporting interoceptive awareness. \emph{Nature Neuroscience}, \emph{7}(2), 189--195.

\leavevmode\vadjust pre{\hypertarget{ref-damasio1996somatic}{}}%
Damasio, A. R. (1996). The somatic marker hypothesis and the possible functions of the prefrontal cortex. \emph{Philosophical Transactions of the Royal Society of London. Series B: Biological Sciences}, \emph{351}(1346), 1413--1420.

\leavevmode\vadjust pre{\hypertarget{ref-desmedt2018heartbeat}{}}%
Desmedt, O., Luminet, O., \& Corneille, O. (2018). The heartbeat counting task largely involves non-interoceptive processes: Evidence from both the original and an adapted counting task. \emph{Biological Psychology}, \emph{138}, 185--188.

\leavevmode\vadjust pre{\hypertarget{ref-desmedt2022does}{}}%
Desmedt, O., Van Den Houte, M., Walentynowicz, M., Dekeyser, S., Luminet, O., \& Corneille, O. (2022). How does heartbeat counting task performance relate to theoretically-relevant mental health outcomes? A meta-analysis. \emph{Collabra: Psychology}, \emph{8}(1), 33271.

\leavevmode\vadjust pre{\hypertarget{ref-ditto2006visceral}{}}%
Ditto, P. H., Pizarro, D. A., Epstein, E. B., Jacobson, J. A., \& MacDonald, T. K. (2006). Visceral influences on risk-taking behavior. \emph{Journal of Behavioral Decision Making}, \emph{19}(2), 99--113.

\leavevmode\vadjust pre{\hypertarget{ref-domschke2010interoceptive}{}}%
Domschke, K., Stevens, S., Pfleiderer, B., \& Gerlach, A. L. (2010). Interoceptive sensitivity in anxiety and anxiety disorders: An overview and integration of neurobiological findings. \emph{Clinical Psychology Review}, \emph{30}(1), 1--11.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,mask,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\linenumbers
\usepackage{csquotes}
\usepackage[labelfont=bf, font={color=gray,small}]{caption}
\usepackage{float}
\usepackage[document]{ragged2e}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={The Heart can Lie: A Preliminary Investigation of the Role of Interoception and Theory of Mind in Deception},
  pdflang={en-EN},
  pdfkeywords={Deception; Interoception; Theory of Mind; Polygraph; Lying Ability},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\section{Methods}\label{methods}

\subsection{Participants}\label{participants}

@@

\subsection{Measures}\label{measures}

\subsubsection{Theory of Mind (ToM)}\label{theory-of-mind-tom}

@@

\subsubsection{Interoception}\label{interoception}

@@

\subsubsection{Deception}\label{deception}

@@

\subsection{Procedure}\label{procedure}

@@

\subsection{Data Analysis}\label{data-analysis}

@@

\section{Results}\label{results}

\subsection{Manipulation Check}\label{manipulation-check}

@@

\subsection{Feature Reduction}\label{feature-reduction}

@@

\subsection{Theory of Mind}\label{theory-of-mind}

@@

\subsection{Interoception}\label{interoception-1}

@@

\section{Discussion}\label{discussion}

The present study examined the contribution of ToM and interoception on our ability to lie using a directed lie paradigm with two conditions (``Interrogation'' and ``Polygraph'') designed to enhance each of the two mechanisms. Interestingly, we found that when participants' responses were perceived to be evaluated by a person (the interrogation condition), instead of (fake) lie detection machine (the polygraph condition), their response time for both lies and truths were faster, and their heart rate was elevated. Although the condition did not impact the subjective confidence that participants had in their answers, the pattern of results suggests that believing one's response is being evaluated by a person, instead of a machine, could induce more fear, consequently speeding up the response and increasing the physiological arousal (Aylward et al., 2017). Alternatively, the slower response in the polygraph condition could be explained by the established attentional switching hypothesis, which posits that an increase in attention towards internal signals and managing one's emotional reaction would confer less cognitive resources available, thereby resulting in individuals taking a longer time to respond (Arnold et al., 2019; Hanania \& Smith, 2010). While the impacts of external settings on individuals' responses warrant further investigation, the results highlight how physiological responses can be easily confounded by other factors (e.g., presence or absence of a stimuli), independent of whether one is lying or telling the truth. By extension, our study concurs with the controversial discourse surrounding the use of physiological measures in deception research (Oviatt et al., 2018; Rosky, 2013).

Furthermore, our results suggest that higher ToM abilities were related to slower and less confident lies, but only in the polygraph condition. While previous bodies of work have reported mixed findings regarding the association between interoception and ToM (Chiou \& Lee, 2013; Gendolla \& Wicklund, 2009; Scaffidi Abbate et al., 2016; Wundrack \& Specht, 2023), our results suggest the two are negatively linked. One possible interpretation of our findings is that people with stronger ToM abilities by default rely more on their social skills and altercentric inference when lying (i.e., they focus on - and try to read - the other person). When that mechanism is unavailable or unsuited (e.g., when there is no person to lie to - but a ``machine'' in our case), their corresponding lying ability decreases. However, in light of the current field of mixed findings relating interoception and ToM (Canino et al., 2022; Gao et al., 2019; Miller, 2015; Shah et al., 2017), future studies are necessary to investigate the interaction of these mechanisms in different social contexts.

We also found that interoceptive abilities (as indicated by the composite interoception scores) are correlated with a higher confidence in one's lies in the polygraph condition, a condition in which the attention towards internal reactions is fostered. Indeed, this is in line with previous studies that found individuals with low interoception were more averse to risk when reputational stakes were high, telling fewer egoistical lies (Vabba et al., 2022). In fact, Vabba et al. (2022) further reported that people with high interoception abilities were less likely to differ in risk-taking tendencies, telling the same number of lies regardless of the social stakes. Consistent with our results, Mohr et al. (2023) found that individuals with high interoceptive accuracy were more likely to make egocentric decisions. However, in contrast to previous studies (Füstös et al., 2013; Owens et al., 2018; Pinna \& Edwards, 2020; Pollatos et al., 2007), we did not find any significant relationship between individuals' interoception scores and their heart rate changes during their answers. This points toward a predominantly meta-cognitive effect without necessarily an actual bodily regulation (i.e., participants with good interoception feel that their lies are more convincing, but do not actively attenuate their bodily reactions).

@@

Additionally to the limitations pertaining to the measure of lying ability, some also concern the measure of the predictor constructs, namely ToM and interoception. While we tried to include a behavioral task as well as a subjective questionnaire for each, it has to be underlined that they are notoriously difficult concepts to measure. In particular, objective interoceptive accuracy was assessed using the Heartbeat Counting Task (HCT). While the HCT used to be considered as a gold standard and remains one of the most commonly used measures (Desmedt et al., 2022), concerns regarding its validity have been increasingly highlighted in several studies as more research efforts are invested into developing novel interoception tasks (Brener \& Ring, 2016; Desmedt et al., 2018, 2022; Legrand et al., 2022; Plans et al., 2021; Ponzo et al., 2021). Future works should further examine the relationship between interoception and lying ability using measures with better psychometric properties.

@@

\section{Data Availability}\label{data-availability}

@@

\section{Conflict of Interest Statement}\label{conflict-of-interest-statement}

@@

\section{Acknowledgements}\label{acknowledgements}

@@

\section{References}\label{references}

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-abe2011brain}
Abe, N. (2011). How the brain shapes deception: An integrated review of the literature. \emph{The Neuroscientist}, \emph{17}(5), 560--574.

\bibitem[\citeproctext]{ref-abe2007deceiving}
Abe, N., Suzuki, M., Mori, E., Itoh, M., \& Fujii, T. (2007). Deceiving others: Distinct neural responses of the prefrontal cortex and amygdala in simple fabrication and deception with social interactions. \emph{Journal of Cognitive Neuroscience}, \emph{19}(2), 287--295.

\bibitem[\citeproctext]{ref-arnold2019interoception}
Arnold, A. J., Winkielman, P., \& Dobkins, K. (2019). Interoception and social connection. \emph{Frontiers in Psychology}, \emph{10}, 2589.

\bibitem[\citeproctext]{ref-aylward2017impact}
Aylward, J., Valton, V., Goer, F., Mkrtchian, A., Lally, N., Peters, S., Limbachya, T., \& Robinson, O. J. (2017). The impact of induced anxiety on affective response inhibition. \emph{Royal Society Open Science}, \emph{4}(6), 170084.

\bibitem[\citeproctext]{ref-baron1997mindblindness}
Baron-Cohen, S. (1997). \emph{Mindblindness: An essay on autism and theory of mind}. MIT press.

\bibitem[\citeproctext]{ref-barrett2015interoceptive}
Barrett, L. F., \& Simmons, W. K. (2015). Interoceptive predictions in the brain. \emph{Nature Reviews Neuroscience}, \emph{16}(7), 419--429.

\bibitem[\citeproctext]{ref-battista2021role}
Battista, F., Otgaar, H., Mangiulli, I., \& Curci, A. (2021). The role of executive functions in the effects of lying on memory. \emph{Acta Psychologica}, \emph{215}, 103295.

\bibitem[\citeproctext]{ref-baumgartner2013honest}
Baumgartner, T., Gianotti, L. R., \& Knoch, D. (2013). Who is honest and why: Baseline activation in anterior insula predicts inter-individual differences in deceptive behavior. \emph{Biological Psychology}, \emph{94}(1), 192--197.

\bibitem[\citeproctext]{ref-beaudoin2020systematic}
Beaudoin, C., Leblanc, É., Gagner, C., \& Beauchamp, M. H. (2020). Systematic review and inventory of theory of mind measures for young children. \emph{Frontiers in Psychology}, \emph{10}, 2905.

\bibitem[\citeproctext]{ref-bond2008individual}
Bond Jr, C. F., \& DePaulo, B. M. (2008). Individual differences in judging deception: Accuracy and bias. \emph{Psychological Bulletin}, \emph{134}(4), 477.

\bibitem[\citeproctext]{ref-bora2017meta}
Bora, E., \& Yener, G. G. (2017). Meta-analysis of social cognition in mild cognitive impairment. \emph{Journal of Geriatric Psychiatry and Neurology}, \emph{30}(4), 206--213.

\bibitem[\citeproctext]{ref-brener2016towards}
Brener, J., \& Ring, C. (2016). Towards a psychophysics of interoceptive processes: The measurement of heartbeat detection. \emph{Philosophical Transactions of the Royal Society B: Biological Sciences}, \emph{371}(1708), 20160015.

\bibitem[\citeproctext]{ref-burgoon1994interpersonal}
Burgoon, J. K., \& Buller, D. B. (1994). Interpersonal deception: III. Effects of deceit on perceived communication and nonverbal behavior dynamics. \emph{Journal of Nonverbal Behavior}, \emph{18}(2), 155--184.

\bibitem[\citeproctext]{ref-Burkner2017}
Bürkner, P.-C. (2017). {brms}: An {R} package for {Bayesian} multilevel models using {Stan}. \emph{Journal of Statistical Software}, \emph{80}(1), 1--28. \url{https://doi.org/10.18637/jss.v080.i01}

\bibitem[\citeproctext]{ref-canino2022embodiment}
Canino, S., Raimo, S., Boccia, M., Di Vita, A., \& Palermo, L. (2022). On the embodiment of social cognition skills: The inner and outer body processing differently contributes to the affective and cognitive theory of mind. \emph{Brain Sciences}, \emph{12}(11), 1423.

\bibitem[\citeproctext]{ref-chen2021emerging}
Chen, W. G., Schloesser, D., Arensdorf, A. M., Simmons, J. M., Cui, C., Valentino, R., Gnadt, J. W., Nielsen, L., Hillaire-Clarke, C. S., Spruance, V., et al. (2021). The emerging science of interoception: Sensing, integrating, interpreting, and regulating signals within the self. \emph{Trends in Neurosciences}, \emph{44}(1), 3--16.

\bibitem[\citeproctext]{ref-chiou2013enactment}
Chiou, W.-B., \& Lee, C.-C. (2013). Enactment of one-to-many communication may induce self-focused attention that leads to diminished perspective taking: The case of facebook. \emph{Judgment and Decision Making}, \emph{8}(3), 372--380.

\bibitem[\citeproctext]{ref-connell2018interoception}
Connell, L., Lynott, D., \& Banks, B. (2018). Interoception: The forgotten modality in perceptual grounding of abstract and concrete concepts. \emph{Philosophical Transactions of the Royal Society B: Biological Sciences}, \emph{373}(1752), 20170143.

\bibitem[\citeproctext]{ref-craig2009you}
Craig, A. D. (2009). How do you feel---now? The anterior insula and human awareness. \emph{Nature Reviews Neuroscience}, \emph{10}(1), 59--70.

\bibitem[\citeproctext]{ref-critchley2004neural}
Critchley, H. D., Wiens, S., Rotshtein, P., Öhman, A., \& Dolan, R. J. (2004). Neural systems supporting interoceptive awareness. \emph{Nature Neuroscience}, \emph{7}(2), 189--195.

\bibitem[\citeproctext]{ref-damasio1996somatic}
Damasio, A. R. (1996). The somatic marker hypothesis and the possible functions of the prefrontal cortex. \emph{Philosophical Transactions of the Royal Society of London. Series B: Biological Sciences}, \emph{351}(1346), 1413--1420.

\bibitem[\citeproctext]{ref-desmedt2018heartbeat}
Desmedt, O., Luminet, O., \& Corneille, O. (2018). The heartbeat counting task largely involves non-interoceptive processes: Evidence from both the original and an adapted counting task. \emph{Biological Psychology}, \emph{138}, 185--188.

\bibitem[\citeproctext]{ref-desmedt2022does}
Desmedt, O., Van Den Houte, M., Walentynowicz, M., Dekeyser, S., Luminet, O., \& Corneille, O. (2022). How does heartbeat counting task performance relate to theoretically-relevant mental health outcomes? A meta-analysis. \emph{Collabra: Psychology}, \emph{8}(1), 33271.

\bibitem[\citeproctext]{ref-ditto2006visceral}
Ditto, P. H., Pizarro, D. A., Epstein, E. B., Jacobson, J. A., \& MacDonald, T. K. (2006). Visceral influences on risk-taking behavior. \emph{Journal of Behavioral Decision Making}, \emph{19}(2), 99--113.

\bibitem[\citeproctext]{ref-dunn2010listening}
Dunn, B. D., Galton, H. C., Morgan, R., Evans, D., Oliver, C., Meyer, M., Cusack, R., Lawrence, A. D., \& Dalgleish, T. (2010). Listening to your heart: How interoception shapes emotion experience and intuitive decision making. \emph{Psychological Science}, \emph{21}(12), 1835--1844.
